High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-04 paper/boneyard.tex
working on discussion, added summaries of a bunch of papers

>>>>

\item Discoveries:
    \begin{itemize}
      \item Dimensionality reduction and graph theoretic approaches
        give different insights into the data and identify different
        patterns as being relevant to cognition (different peak
        orders).
        \item An insight common to both approaches is that high-order
          (greater than first order) dynamic correlations are
          informative about ongoing high-level cognitive processing.
          As the level of cognitive processing decreases, cognition is
          reflected by lower-order correlations.
        \item Correlations at different orders are also associated
          with different networks of brain regions.  However, which
          networks reflect which types of interactions depends on the
          current task.  In general, lower order correlations during
          auditory listening reflect processing of low-level
          (auditory) features; mid-order correlations reflect speech
          and linguistic processing; higher-order correlations reflect
          across-sensory integration (e.g. ties to visual areas) and
          cognitive control areas.  This hierarchy dissolves during
          lower-order cognitive processing.
      \end{itemize}

      %%%%%%%

Based on prior work ~\citep{DemeEtal19} and following the direction of the field ~\citep{Turk13} we think our thoughts might be encoded in
dynamic network patterns, and possibly higher order network
patterns (Fig.~\ref{fig:direction_of_field}). We sought to test this hypothesis by developing an approach
to inferring high-order network dynamics from timeseries data.

One challenge in studying dynamic interactions is the
computational resources required to calculate higher-order correlations. 
We developed a computationally tractable model of network dynamics (Fig.~\ref{fig:methods}) that takes in a feature
timeseries and outputs approximated first-order dynamics (i.e.,
dynamic functional correlations), second-order dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).

We first validated our model using synthetic data, and explored how
recovery varied with different underlying data structures and kernels.   We then 
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio
recording of a story, as well as scrambled versions of the same story
(where the scrambling was applied at different temporal scales).  We
trained classifiers to take the output of the model and decode the
timepoint in the story (or scrambled story) that the participants were
listening to. We found that, during the intact listening condition in the
experiment, classifiers that incorporated higher-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  a.\&d.).  By contrast, these
higher-order correlations were not necessary to support decoding the other
listening conditions and (minimally
above chance) during a control rest condition.  This suggests
that the cognitive processing that supported the most cogntively rich listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best when incorporating
higher-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique.  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions. Exploring this method
  with other data collection technique will be important to
  disentangle this question.

  \subsection*{Concluding remarks}

How can we better understand how brain patterns change over
time? How can we quantify the potential network dynamics that might be
driving these changes? One way to judge the techniques of the future is
to look at the trajectory of the fMRI field so far has taken so far
(Fig.~\ref{fig:methods}).  The field started with 
univariate activation, measuring the average activity for each voxel.
Analyses of multivariate activation followed, looking at spatial patterns of
activity over voxels. Next, correlations of activity were explored, first
with measures like resting connectivity that take temporal correlation
between a seed voxel and all other voxels then with full connectivty
that measure all pairwise correlations.  Additionally, this path of increasing
complexity also moved from static to dynamic measurements.  One
logical next step in this trajectory would be dynamic higher-order
correlations. We have created a method 
to support these calculations by scalably approximating dynamic higher-order
correlations.

### 2019-09-04 paper/main.tex
added discussion figure
working on discussion, added summaries of a bunch of papers

\textbf{JRM STOPPED HERE}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

\begin{itemize}
  \item Methods advances: kernel-based dynamic correlations, extension to
    dynamic ISFC, efficient method for estimating high-order dynamic
    correlations, identifying robust results by averaging
  \item Discoveries:
    \begin{itemize}
      \item Dimensionality reduction and graph theoretic approaches
        give different insights into the data and identify different
        patterns as being relevant to cognition (different peak
        orders).
        \item An insight common to both approaches is that high-order
          (greater than first order) dynamic correlations are
          informative about ongoing high-level cognitive processing.
          As the level of cognitive processing decreases, cognition is
          reflected by lower-order correlations.
        \item Correlations at different orders are also associated
          with different networks of brain regions.  However, which
          networks reflect which types of interactions depends on the
          current task.  In general, lower order correlations during
          auditory listening reflect processing of low-level
          (auditory) features; mid-order correlations reflect speech
          and linguistic processing; higher-order correlations reflect
          across-sensory integration (e.g. ties to visual areas) and
          cognitive control areas.  This hierarchy dissolves during
          lower-order cognitive processing.
      \end{itemize}
  \end{itemize}

Based on prior work ~\citep{DemeEtal19} and following the direction of the field ~\citep{Turk13} we think our thoughts might be encoded in
dynamic network patterns, and possibly higher order network
patterns (Fig.~\ref{fig:direction_of_field}). We sought to test this hypothesis by developing an approach
to inferring high-order network dynamics from timeseries data.

One challenge in studying dynamic interactions is the
computational resources required to calculate higher-order correlations. 
We developed a computationally tractable model of network dynamics (Fig.~\ref{fig:methods}) that takes in a feature
timeseries and outputs approximated first-order dynamics (i.e.,
dynamic functional correlations), second-order dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).

We first validated our model using synthetic data, and explored how
recovery varied with different underlying data structures and kernels.   We then 
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio
recording of a story, as well as scrambled versions of the same story
(where the scrambling was applied at different temporal scales).  We
trained classifiers to take the output of the model and decode the
timepoint in the story (or scrambled story) that the participants were
listening to. We found that, during the intact listening condition in the
experiment, classifiers that incorporated higher-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  a.\&d.).  By contrast, these
higher-order correlations were not necessary to support decoding the other
listening conditions and (minimally
above chance) during a control rest condition.  This suggests
that the cognitive processing that supported the most cogntively rich listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best when incorporating
higher-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique.  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions. Exploring this method
  with other data collection technique will be important to
  disentangle this question.

  \subsection*{Concluding remarks}

How can we better understand how brain patterns change over
time? How can we quantify the potential network dynamics that might be
driving these changes? One way to judge the techniques of the future is
to look at the trajectory of the fMRI field so far has taken so far
(Fig.~\ref{fig:methods}).  The field started with 
univariate activation, measuring the average activity for each voxel.
Analyses of multivariate activation followed, looking at spatial patterns of
activity over voxels. Next, correlations of activity were explored, first
with measures like resting connectivity that take temporal correlation
between a seed voxel and all other voxels then with full connectivty
that measure all pairwise correlations.  Additionally, this path of increasing
complexity also moved from static to dynamic measurements.  One
logical next step in this trajectory would be dynamic higher-order
correlations. We have created a method 
to support these calculations by scalably approximating dynamic higher-order
correlations.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
We tested the hypothesis that high-level cognition is supported by
high-order brain network dynamics.  We examined high-order network
dynamics in functional neuroimaging data collected during a story
listening experiment.  When participants listened to an auditory
recording of the story, participants exhibited similar high-order
brain network dynamics.  By contrast, when participants instead
listened to temporally scrambled recordings of the story, only
lower-order brain network dynamics were similar across participants.
Our results indicate that higher orders of network interactions
support higher-level aspects of cognitive processing
(Fig.\ref{fig:discussion}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/discussion}
  \caption{\textbf{Proposed high-order network dynamics underlying
      high-level cognition.}
    Higher orders of network interactions support higher-level aspects
    of cognitive processing.  When tasks evoke richer, deeper, and/or
    higher-level processing, this is reflected in higher-order network
    interactions.}
  \label{fig:discussion}
\end{figure}

Identifying higher-order network dynamics associated with high-level
cognition required several important methods advances.  First, we used
kernel-based dynamic correlations to extended the notion of (static)
inter-subject functional connectivity~\citep{SimoEtal16} to a dynamic
measure of inter-subject functional connectivity (DISFC) that does not
rely on sliding windows, and that may be computed at individual
timepoints.  This allowed us to precisely characterize stimulus-evoked
network dynamics that were similar across individuals.  Second, we
developed a computational framework for efficiently and scalably estimating
high-order dynamic correlations.  Our approach uses dimensionality
reduction algorithms and graph measures to obtain low-dimensional
embeddings of patterns of network dynamics.  Third, we developed an
analysis framework for identifying robust decoding results by carrying
out our analyses using a range of parameter values and then
identifying which results were robust to specific parameter choices.

Relevant literature:
\begin{itemize}
\item \cite{SoloEtal19}: more complex concepts involve more complex network
dynamics

\item \cite{DemeEtal19}: human consciousness is supported by dynamic
complex patterns of brain signal coordination.  studied the brains of
healthy vs minimally conscious patients; conscious patients displayed
richer and more structured patterns of network interactions

\item \cite{MackEtal17}: richer task structure is supported by more complex (higher
rank) brain patterns in the prefrontal cortex

\item \cite{LuriEtal19}: review of time-varying connectivity

\item \cite{FongEtal18}: higher variability in dynamic functional connectivity
predicts worse attention.  also, dymamic functional connectivity
improves attention predictions over static functional connectivity.

\item \cite{ParkEtal17}: dynamic causal modelling: compute causal interaction
dynamics in multivariate timeseries data.  important to consider
causation in interpreting the direction of information flow reflected
in the patterns we identified through our analyses.

\item \cite{PretEtal17}: another review of time varying connectivity

\item \cite{RoyEtal19}: memories are distributed across functionally connected
brain areas

\item \cite{FallEtal19}: the timescale of spontaneous resting state
  fMRI increases with structural connectivity strength (in mouse and
  human brains)

\item \cite{LiegEtal19}: dynamic resting state functional connectivity best captures
task-based phenotypes (processing speed, fluid intelligence), whereas
static and dynamic functional connectivity equally capture
self-reported measures like loneliness and life satisfaction.  suggest
that different timescales of connectivity dynamics reflect different
aspects of behavior.

\item \cite{ZhenEtal19}: characterize brain connectectomes at different geometric
scales

\item \cite{McInJirs19}: neural processes reflect system interactions that occur on
relatively low-dimensional manifolds, thereby constraining possible
functional configurations.  (motivation for dimensionality reduction
approach?)

\item \cite{TokeSomm19}: another approach to reducing dimensionality of network
interactions, using spectral clustering

\item \cite{ZouEtal19}: another review of complex network dynamics in timeseries
data

\item \cite{ShapEtal19}: state-based analyses of network dynamics using hidden
semi-Markov models.  could be a useful direction for extended our
work; for example, we could define kernel functions to only integrate
within an ``event.''  This could potentially improve resolution of
event-specific network dynamics.

\item \cite{ColeEtal19}: potential artifacts or spurious insights that affect
standard functional connectivity analyses (not sure this applies)

\item \cite{RichEtal18}: simulations for benchmarking time-varying connectivity
methods (should we use this? seems useful! could potentially replace
synthetic data section)

\item \cite{ChanGlob09}

\item \cite{SizeEtal18}

\item \cite{ReimEtal17}

\item \cite{GonzEtal19}

\item \cite{BetzEtal19}

\end{itemize}
